\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-13 paper.tex

This view of ARGs,
decoupled from generative models and
without the hard requirement
of complete precision on all historical events, may clarify inference goals and improve
methods for evaluation.
In most cases,
ARG inference is evaluated by simulating data from a known ground truth ARG,
and comparing this to the inferred version via pairwise comparison of
local trees along the genome, for example
using tree distance
metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping},
as described by \citet{kuhner2015assessing}.
% While this provides valuable insights into the method's performance,
% it is not clear how well these tree distance metrics reflect
% performance on real data.
In comparing tree-by-tree along the genome, the effects of recombination
are only incorporated in an indirect manner through the correlations
between the local trees, rather than through directly taking into account
the persistence of nodes and edges across multiple trees.
Tree distance metrics usually have $O(n^2)$ time complexity and therefore cannot be
applied to the very large sample sizes currently of interest.
A recent trend has been to move away from such tree distance-based approaches and to examine more properties of the inferred ARGs.
\citet{brandt2021evaluation} focussed on the
distributions of pairwise MRCA times; \citet{deng2021distribution} derived an approximation to the distribution of the
waiting distances between local trees; \citet{ignatieva2023distribution}
derived the distribution of the genomic span of an edge and that of a
clade of samples. In each case, simulation studies demonstrated
substantial differences between these quantities in simulated and reconstructed ARGs,
that were not captured using tree-by-tree comparisons.
% Not sure this is the right place for this sentence
These evaluations have been based on
ground truth data from highly idealised simulations, with sample sizes limited to at most a few thousand (typically much fewer).
The effects of the richness of real data
on modern biobank-scale or multi-generational pedigree datasets
are almost entirely unknown (beyond studies using very simplistic error models).
The development of ARG evaluation metrics that take into account more of the
global topology and can be applied to large ARGs would be a
valuable and timely addition to the field.

@@

Methods to sample ARGs generate a ``cloud'' of estimates
\citep{kuhner2017consensus} develop

>>>>

This view of ARGs,
decoupled from generative models and
without the hard requirement
of complete precision on all historical events, may clarify inference goals and improve
methods for evaluation.
In most cases,
ARG inference is evaluated by simulating data from a known ground truth ARG,
and comparing this to the inferred version via pairwise comparison of
local trees along the genome, for example
using tree distance
metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping},
as described by \citet{kuhner2015assessing}.
% While this provides valuable insights into the method's performance,
% it is not clear how well these tree distance metrics reflect
% performance on real data.
In comparing tree-by-tree along the genome, the effects of recombination
are only incorporated in an indirect manner through the correlations
between the local trees, rather than through directly taking into account
the persistence of nodes and edges across multiple trees.
The performance of tree distance metrics varies by
application~\citep{kuhner2015assessing}, and the correct approach
to handling subtleties such as polytomies
is an open question~\citep{kelleher2019inferring,zhang2023biobank}.
% Is this true? These guys claim RF takes linear time:
% https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-020-07011-0
Tree distance metrics usually have at least $O(n^2)$ time complexity
and therefore cannot be
applied to the very large sample sizes currently of interest.
A recent trend has been to move away from such
tree distance-based approaches and to examine more properties of the inferred ARGs.
\citet{brandt2021evaluation} focussed on the
distributions of pairwise MRCA times;
\citet{deng2021distribution} derived an approximation to the distribution of the
waiting distances between local trees;
\citet{ignatieva2023distribution}
derived the distribution of the genomic span of an edge and that of a
clade of samples. In each case, simulation studies demonstrated
substantial differences between these quantities in simulated and reconstructed ARGs,
that were not captured using tree-by-tree comparisons.
% Not sure this is the right place for this sentence
These evaluations have been based on
ground truth data from highly idealised simulations, with sample sizes limited to at most a few thousand (typically much fewer).
The effects of the richness of real data
on modern biobank-scale or multi-generational pedigree datasets
are almost entirely unknown (beyond studies using very simplistic error models).
The development of ARG evaluation metrics that take into account more of the
global topology and can be applied to large ARGs would be a
valuable and timely addition to the field.

@@

Methods to sample ARGs generate a ``cloud'' of estimates, and
\citep{kuhner2017consensus} provide an approach to generate a
set of consensus breakpoints and local trees from
such a cloud.
The approach is based on examining the recombination breakpoints
in all of the input ARGS, and including those that are
in at least $k$ of the input ARGs (with some additional
filtering criteria) in the output.
Within the resulting intervals, a consensus
local tree is then generated using standard phylogenetic methods.
